A digital filter designing tool for the Hankel and Fourier transforms in potential-, diffusive-, and wavefield modeling

### 2017-12-24 ManuscriptGeophysics2018.tex
Draft V2 of manuscript

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
%
% (1) PRINCIPAL OBJECTIVES AND SCOPE OF THE WORK
  The open-source code \texttt{fdesign} makes it possible to design digital
  linear filters for the Hankel and Fourier transforms used in potential-,
  diffusive-, and wavefield modeling. Digital filters can be derived for any
  electromagnetic method, such as methods in the diffusive limits (DC, CSEM) as
  well as methods using higher frequency content (GPR, acoustic and elastic
  wavefields).
%
% (2) METHODOLOGY
  The direct matrix inversion method is used for the derivation of the filter
  values, and a brute-force inversion is carried out over the defined spacing
  and shifting values of the filter basis. Included or user-provided
  theoretical transform pairs are used for the inversion. Alternatively, one
  can provide 1D subsurface models that will be computed with a precise
  quadrature method using the EM modeler \texttt{empymod} to generate
  numerical transform pairs.
%
% (3) RESULTS
  The comparison of the presented 201\,pt filter with previously presented
  filters shows that it performs better for some standard CSEM cases. The
  derivation of longer 2001\,pt filter for a GPR example with 250\,MHz centre
  frequency proves that the filter method works also for wave phenomena, not
  only for diffusive EM fields.
%
% (4) CONCLUSIONS
  The presented algorithm provides a tool to create problem specific digital
  filters. Such purpose-built filters can be made shorter and can speed up
  consecutive  potential-, diffusive-, and wavefield inversions.
%
\end{abstract}

@@

\cite{PhD.70.Ghosh} derived the filter coefficients in the spectral domain by
dividing the output spectrum by the input spectrum followed by an inverse
Fourier transform. Improvements to the determination of filter coefficients
were provided by \cite{EXG.75.ONeill, GEO.77.Nyman, GEO.82.Das}, or
specifically for the Fourier transform by \cite{GP.86.Nissen}. A direct
integration method was used by \cite{GP.76.Bichara, GP.78.Bernabini}.
\cite{GP.79.Koefoed} proposed a Wiener-Hopf least-squares method, which was
further improved by many authors \citep{GP.82.Guptasarma, GEO.1982.Murakami,
GP.97.Guptasarma}. \cite{GP.07.Kong} proposes a direct matrix inversion method
to solve the convolution equation, which requires only the input and output
sample values. To evaluate the filters he defines the criteria of a good filter
as one that recovers small or weak wavefields. This method was also used by
\cite{GEO.09.Key, GEO.12.Key}. Most authors publish filters for the Hankel
transform with $J_0$ and $J_1$ Bessel functions (or $J_{-1/2}$, $J_{1/2}$ if
applied to the Fourier sine/cosine transform), as all higher Bessel functions
can be rewritten to only use these two. \cite{GP.94.Mohsen} is one of the rare
cases which provides $J_2$ filter weights.

@@

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-penetrating radar (GPR). Generally it was even
thought that the filter method works only in the diffusive limit
\citep[e.g.,][]{GEO.15.Hunziker}. Also, there is no open-source filter
designing tool readily available, they are available as a described method
in articles or as code upon request by the author, if at all. The presented
algorithm \texttt{fdesign} tries to fill this gap by providing the tools to
design general or purpose-built filters using the direct matrix inversion
method. After a brief overview of the methodology, the theory, and the code we
show examples of its usage, presenting some new filters for both, CSEM and GPR
data. The algorithm and many more examples of its usage can be found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

The main points of the method and some differences between it and the methods
it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force).
  \item \cite{GP.07.Kong, GEO.12.Key} optimize a filter for the $J_0$ filter,
    and then use the obtained shift and spacing value to calculate the $J_1$
    values. The presented algorithm can optimize either $J_0$, $J_1$, or both
    in the optimization process. Five transform-pairs for each $J_0$ and $J_1$,
    and three transform pairs for each sine and cosine of often used functions
    are included in the routine, but any other transform pair can be provided
    as input.
  \item More complex or more specific models can be provided instead of
    theoretical transform pairs by using the modeler \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In the presented algorithm you can define
    a relative error level which defines up to what error an obtained result is
    good or not. The presented algorithm can also minimize the amplitude, but
    it has additionally another mode, where you maximise the abscissae $r$ of
    the right-hand-side of the transform pair instead of minimising the
    amplitude. The two modes yield the same result in the general, simple case
    of fast decaying transform pairs. However, if you use complex models for
    the design, and specifically high frequencies, then maximising $r$ will
    yield much better and more consistent results.
  \item The algorithm allows to run under-, equal-, and over-determined
    systems.
  \item the real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

@@

Most of the articles mentioned in the review have detailed derivations of the
digital filter method. In this article we focus on the algorithm, and summarize
the theory only very briefly by following \cite{GEO.12.Key} (equations
(2)-(6)). In electromagnetics we often have to evaluate integrals of the form
%
\begin{equation}
  F(r) = \int^\infty_0 f(l)K(l r)\mr{d}l \ ,
  \label{eq:HankelInt}
\end{equation}
%
where $l$ and $r$ denote left-hand-side and right-hand-side evaluation values,
and $K$ is the kernel function. In the specific case of the Hankel transform
$l$ corresponds to wavenumber, $r$ to offset, and $K$ to Bessel functions; in
the case of the Fourier transform $l$ corresponds to frequency, $r$ to time,
and $K$ to sine or cosine functions. In both cases it is an infinite integral
which numerical integration is very time-consuming because of the slow decay
of the kernel function and its oscillatory behaviour.

By substituting $r = e^x$ and $l = e^{-y}$ we get
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
\begin{equation}
  b_n = \exp\left\{\Delta(-N/2+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
From equation~\ref{eq:filtapprox} it can be seen that the filter method
requires $N$ evaluations at each $r$. So to calculate for instance the
frequency domain result for 100 offsets with a 201\,pt filter requires 20'100
evaluations in the wavenumber domain. This is why the DLF often uses
interpolation to minimise the required evaluations, either in the right-hand
side or in the left-hand side.

@@

\subsection{Design}
Figure \ref{fig:201} shows the solution spaces of four consecutive inversion
runs, where each run is a focus on a subsection of the previous run, indicated
by the red square.
%
\plot*{201}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%
The filter length in this case is $N=201$, and the theoretical transform pairs
is (as given, e.g., in \cite{USGS.75.Anderson})
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair for the inversion, and a different pair to get the minimum amplitude or the maximum $r$. In
this case the same transform pair was used for both the inversion and the check
of quality. The acceptable relative error was set to 1\,\%.
The rhs evaluation parameter for the inversion was set to $r_\mr{def} = (1, 1,
2)$, which means that an over-determined system was evaluated with $2N$
equations, where $r$ was logarithmically spaced from $\log_{10}(1/\max(b)) - 1$
to $\log_{10}(1/\min(b)) + 1$, where $b$ are the filter abscissae as given in
equation~\ref{eq:base}.

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often you will come across a good filter by shear
luck of choosing the right starting parameters. Each of the input parameters
has an effect to the outcome. It depends a lot on the transform pairs
$f_\mr{I;C}$, and functions that decay rapidly are generally better, as noted
by earlier authors \citep[e.g.][]{USGS.75.Anderson}. It also depends on the
filter length, and obviously heavily on the spacing and shift values, as this
is what we invert for. Another important point is how you define the right hand
evaluation points of the inversion ($r_\mr{def}$). Evaluating corresponding
transform pairs separately or jointly also leads to different filter
coefficients ($J_0$, $J_1$, or $J_0$ \& $J_1$, or equally sine, cosine, or sine
\& cosine); if the real or the imaginary part is used when complex transform
pairs or \texttt{empymod} is used; and if it is inverted for the minimum
amplitude or for the maximum $r$.

@@

Figures \ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his figure 5. The model consists
of a water layer with $\rho_\mr{w} = 0.3125\,\Omega\,$m of infinite thickness
above a subsurface half-space with $\rho = 1\,\Omega\,$m. The signal of an
$x$-directed electric source 50\,m above the interface is measured at an
$x$-directed electric receiver at the interface, frequency is $f=1\,$Hz.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of figure 5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of 1e-12 and 1e-30, respectively. The
relative error is shown in (b), where the QWE result was taken as \emph{truth}.
For offsets greater than roughly 15.5\,km the relative error becomes
meaningless, as the QWE fails itself; this part is greyed out in the figure.

Figure \ref{fig:keyf5} (a) is the canonical CSEM model as used in
\cite{GEO.12.Key} in his figure 5: Water depth of 2\,km with
$\rho_\mr{w}=0.303\,\Omega\,$m, a background resistivity of
$\rho_\mr{b}=1\,\Omega\,$m, in which a target is embedded of
$\rho_\mr{t}=100\,\Omega\,$m at 1000\,m below the seafloor, 100\,m thick.
Source depth is 1990\,m, receivers are on the seafloor.
%
\plot*{keyf5}{width=.9\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  magnitudes lower than the other filters.}
%
Figure \ref{fig:keyf5} (b) is a land case, with a background resistivity of
$\rho_\mr{b}=10\,\Omega\,$m, in which a target is embedded of
$\rho_\mr{t}=500\,\Omega\,$m at 1000\,m below the surface, 100\,m thick. Source
depth is 0.5\,m, receiver depth is 0.8\,m. In both cases the new filter
\emph{Wer201} has generally a relative error which is magnitudes lower than the
other filters.

@@

\rmk{TODO: Add second GPR result; layers swapped}

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker, GEO.17.Werthmuller}; the model parameters are given in
subplot (a). The filter used for this example for the Hankel transform is a
2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for the
inversion and the check of quality. Vertical source and receiver separation is
1\,m, resistivity $\rho=200\,\Omega\,$m, $\varepsilon_\mr{r}$=10,
$\mu_\mr{r}=1$. For the Fourier transform a 4096\,pt FFT was used with
regularly spaced frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up
to 2048\,MHz for both the calculation with the adaptive quadrature (QUA) and
with the DLF. The frequency result is convolved with a Ricker wavelet with a
center frequency of 250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is
applied. The calculation with the digital filter took under 9 minutes and is
therefore roughly 80 times faster then the adaptive quadrature calculation
which took roughly 11 hours and 27 minutes. Calculating the same model with QWE
took 7 hours and 20 minutes. However, QWE uses in roughly 1/3 of the
calculation internally the adaptive quadrature in this example, see
\cite{GEO.17.Werthmuller}. (Note that DLF was run in parallel using 4 threads
at once, taking effectively only 2 minutes and 10 seconds to calculate.
%EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
%minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.)
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QAGSE from the Fortran
  QUADPACK library) or with (c) a 2001\,pt digital linear filter (b). The
  quadrature took about 11\,h 27\,min to calculate, whereas the filter took
  2\,min 10\,s.}
%

Figure~\ref{fig:ftGPR} shows in (a)-(c) the real part of the frequency domain
results and in (d)-(f) the time domain results for offsets of 0.2\,m, 2.0\,m,
and 3.0\,m.
%
\plot*{ftGPR}{width=\textwidth}{The real-part frequency-domain and the
time-domain responses for offsets of (a) 0.2\,m, (b) 2.0\,m, and (c) 3.0\,m.}
%
These examples show clearly that the filter method can indeed be applied to
high frequency EM modeling and therefore wave phenomena. We are convinced that
with further tests and analysis much better filters could be achieved, and
various concepts could be checked. One approach is to derive a filter for each
frequency band, say one for 1\,MHz--10\,MHz, one for 10\,MHz--100\,MHz, and one
for 100\,MHz--1\,GHz. Another approach could be to derive distinct filters for
$J_0$ and $J_1$, with different spacing and shift values. The first idea would
triple the calculation cost, the second idea double them; however, they both
would still be very fast compared to standard quadrature methods.

@@

The presented, free and open-source algorithm \texttt{fdesign} can be used to
design digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}.
The code is available from GitHub in the add-on repository of \texttt{empymod},
called \texttt{empyscripts}, from version 0.1.2 onwards.

@@

We hope the presented algorithm is useful for at least 3 scenarios:
\begin{enumerate}
  \item Provide a fast method to design problem-specific filters. For bigger
    inversion projects (such as, for instance, massive stochastic CSEM
    inversions) a purpose designed, short filter might save much more time than
    it costs to design it. This could even be integrated into inversion codes,
    as an optional pre-inversion step.
  \item Extend the filter method to new areas, namely higher frequencies.
  \item Raise a new interest for digital linear filters in geophysics, by
    making it very easy for anyone to play around, create their own filters,
    and get a better understanding of it. We are sure there are many great
    filters to be discovered.
\end{enumerate}

\section{Acknowledgment}
D.W. would like to thank his colleague Oleg Gurevich Titov, who introduced him
to linear digital filters in the first place.
% This project would not have been successful without the fruitful discussions
% with Evert Slob and Kerry Key. Many thanks as well to Kerry Key for providing
% me with his Matlab scripts, which served as a starting point and initiated
% this project.

% REFERENCES
\bibliographystyle{seg}
\bibliography{\string~/.dotfiles/LaTeX/dtrRef}

>>>>

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
%
% (1) PRINCIPAL OBJECTIVES AND SCOPE OF THE WORK
  The open-source code \texttt{fdesign} makes it possible to design digital
  linear filters for the Hankel and Fourier transforms used in potential-,
  diffusive-, and wavefield modeling. Digital filters can be derived for any
  electromagnetic method, such as methods in the diffusive limits (DC, CSEM) as
  well as methods using higher frequency content (GPR, acoustic and elastic
  wavefields).
%
% (2) METHODOLOGY
  The direct matrix inversion method is used for the derivation of the filter
  values, and a brute-force inversion is carried out over the defined spacing
  and shifting values of the filter basis. Included or user-provided
  theoretical transform pairs are used for the inversion. Alternatively, one
  can provide 1D subsurface models that will be computed with a precise
  quadrature method using the EM modeler \texttt{empymod} to generate
  numerical transform pairs.
%
% (3) RESULTS
  The comparison of the presented 201\,pt filter with previously presented
  filters shows that it performs better for some standard CSEM cases. The
  derivation of a longer 2001\,pt filter for a GPR example with 250\,MHz centre
  frequency proves that the filter method works also for wave phenomena, not
  only for diffusive EM fields.
%
% (4) CONCLUSIONS
  The presented algorithm provides a tool to create problem specific digital
  filters. Such purpose-built filters can be made shorter and can speed up
  consecutive  potential-, diffusive-, and wavefield inversions.
%
\end{abstract}

@@

\cite{PhD.70.Ghosh} derived the filter coefficients in the spectral domain by
dividing the output spectrum by the input spectrum followed by an inverse
Fourier transform. Improvements to the determination of filter coefficients
were provided by \cite{EXG.75.ONeill, GEO.77.Nyman, GEO.82.Das}, or
specifically for the Fourier transform by \cite{GP.86.Nissen}. A direct
integration method was used by \cite{GP.76.Bichara, GP.78.Bernabini}.
\cite{GP.79.Koefoed} proposed a Wiener-Hopf least-squares method, which was
further improved by many authors \citep{GP.82.Guptasarma, GEO.1982.Murakami,
GP.97.Guptasarma}. \cite{GP.07.Kong} proposed a direct matrix inversion method
to solve the convolution equation, which requires only the input and output
sample values. To evaluate the filters he defines the criteria of a good filter
as one that recovers small or weak diffusive EM fields. This method was also
used by \cite{GEO.09.Key, GEO.12.Key}. Most authors publish filters for the
Hankel transform with $J_0$ and $J_1$ Bessel functions (or $J_{-1/2}$,
$J_{1/2}$ if applied to the Fourier sine/cosine transform), as all higher
Bessel functions can be rewritten to only use these two. \cite{GP.94.Mohsen} is
one of the rare cases which provides $J_2$ filter weights.

@@

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-pe\-ne\-tra\-ting radar (GPR). Generally it
was even thought that the filter method works only in the diffusive limit
\citep[e.g.,][]{GEO.15.Hunziker}. Also, there is no open-source filter
designing tool readily available, they are available as a described method
in articles or as code upon request by the author, if at all. The presented
algorithm \texttt{fdesign} tries to fill this gap by providing the tools to
design general or purpose-built filters using the direct matrix inversion
method. After a brief overview of the methodology, the theory, and the code we
show examples of its usage, presenting some new filters for both, CSEM and GPR
data. The algorithm and many more examples of its usage can be found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

The main points of the method and some differences between it and the methods
it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force), with an optional minimization routine for
    the best outcome of the brute-force step.
  \item \cite{GP.07.Kong, GEO.12.Key} optimize a filter for the $J_0$ filter,
    and then use the obtained shift and spacing value to calculate the $J_1$
    values. The presented algorithm can optimize either $J_0$, $J_1$, or both
    in the optimization process. Five transform-pairs for each $J_0$ and $J_1$,
    and three transform pairs for each sine and cosine of often used functions
    are included in the routine, but any other transform pair can be provided
    as input.
  \item More complex or more specific models can be provided instead of
    theoretical transform pairs by using the modeler \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In the presented algorithm you can define
    a relative error level which defines up to what error an obtained result is
    good or not. The presented algorithm can also minimize the amplitude, but
    it has additionally another mode, where you maximise the abscissae $r$ of
    the right-hand-side of the transform pair instead of minimising the
    amplitude. The two modes yield the same result in the general, simple case
    of fast decaying transform pairs. However, if you use complex models for
    the design, and specifically high frequencies, then maximising $r$ will
    yield much better and more consistent results.
  \item The algorithm allows to run under-, equal-, and over-determined
    systems.
  \item The real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

@@

Most of the articles mentioned in the review have detailed derivations of the
digital filter method. In this article we focus on the algorithm, and summarize
the theory only very briefly by following \cite{GEO.12.Key}. In
electromagnetics we often have to evaluate integrals of the form
%
\begin{equation}
  F(r) = \int^\infty_0 f(l)K(l r)\mr{d}l \ ,
  \label{eq:HankelInt}
\end{equation}
%
where $l$ and $r$ denote left-hand-side and right-hand-side evaluation values,
and $K$ is the kernel function. In the specific case of the Hankel transform
$l$ corresponds to wavenumber, $r$ to offset, and $K$ to Bessel functions; in
the case of the Fourier transform $l$ corresponds to frequency, $r$ to time,
and $K$ to sine or cosine functions. In both cases it is an infinite integral
which numerical integration is very time-consuming because of the slow decay
of the kernel function and its oscillatory behaviour.

By substituting $r = e^x$ and $l = e^{-y}$ we get
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
\begin{equation}
  b_n = \exp\left\{\Delta(-N/2+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
From equation~\ref{eq:filtapprox} it can be seen that the filter method
requires $N$ evaluations at each $r$. To calculate for instance the frequency
domain result for 100 offsets with a 201\,pt filter requires 20'100 evaluations
in the wavenumber domain. This is why the DLF often uses interpolation to
minimise the required evaluations, either in the right-hand side (often called
\emph{lagged convolution DLF}) or in the left-hand side (\emph{splined DLF}).

@@

\subsection{Design}
Figure \ref{fig:201} shows the solution spaces of four consecutive inversion
runs, where each run is a focus on a subsection of the previous run, indicated
by the red square.
%
\plot*{201}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%
The filter length in this case is $N=201$, and the theoretical transform pairs
are (as given, e.g., in \cite{USGS.75.Anderson})
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair for the inversion, and a different pair to get the minimum amplitude or the maximum $r$. In
this case the same transform pair was used for both the inversion and the check
of quality. The acceptable relative error was set to 1\,\%.
The rhs evaluation parameter for the inversion was set to $r_\mr{def} = (1, 1,
2)$, which means that an over-determined system was evaluated with $2N$
equations, where $r$ was logarithmically spaced from $\log_{10}(1/\max(b)) - 1$
to $\log_{10}(1/\min(b)) + 1$, where $b$ are the filter abscissae as given in
equation~\ref{eq:base}.

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often you will come across a good filter by shear
luck of choosing the right starting parameters. Each of the input parameters
has an effect to the outcome. It depends a lot on the transform pairs
$f_\mr{I;C}$, and functions that decay rapidly are generally better, as noted
by earlier authors \citep[e.g.][]{USGS.75.Anderson}. It also depends on the
filter length, and obviously heavily on the spacing and shift values, as this
is what we invert for. Another important point is how you define the right-hand
evaluation points of the inversion ($r_\mr{def}$). Evaluating corresponding
transform pairs separately or jointly also leads to different filter
coefficients ($J_0$, $J_1$, or $J_0$ \& $J_1$, or equally sine, cosine, or sine
\& cosine); also if the real or the imaginary part is used when complex
transform pairs or \texttt{empymod} is used; and if it is inverted for the
minimum amplitude or for the maximum $r$.

@@

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his figure 5. The model consists
of a water layer with $\rho_\mr{w} = 0.3125\,\Omega\,$m of infinite thickness
above a subsurface half-space with $\rho = 1\,\Omega\,$m. The signal of an
$x$-directed electric source 50\,m above the interface is measured at an
$x$-directed electric receiver at the interface, frequency is $f=1\,$Hz.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of figure 5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of 1e-12 and 1e-30, respectively. The
relative error is shown in (b), where the QWE result was taken as \emph{truth}.
For offsets greater than roughly 15.5\,km the relative error becomes
meaningless, as the QWE fails itself; this part is greyed out in the figure.

Figure \ref{fig:keyf5} (a) is the canonical CSEM model as used in
\cite{GEO.12.Key} in his figure 5: Water depth of 2\,km with
$\rho_\mr{w}=0.303\,\Omega\,$m, a background resistivity of
$\rho_\mr{b}=1\,\Omega\,$m, in which a target is embedded of
$\rho_\mr{t}=100\,\Omega\,$m at 1000\,m below the seafloor, 100\,m thick.
Source depth is 1990\,m, receivers are on the seafloor.
%
\plot*{keyf5}{width=.9\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.}
%
Figure \ref{fig:keyf5} (b) is a land case, with a background resistivity
of $\rho_\mr{b}=10\,\Omega\,$m, in which a target is embedded of
$\rho_\mr{t}=500\,\Omega\,$m at 1000\,m below the surface, 100\,m thick. Source
depth is 0.5\,m, receiver depth is 0.8\,m. In both cases the new filter
\emph{Wer201} has generally a relative error which is orders of magnitude
lower than the other filters. The relative error of the real part is given
in black, whereas the relative error of the imaginary part is given in grey.
It is interesting to note that the real and imaginary parts have very similar
errors in \emph{Wer201}, \emph{Key201}, and \emph{And801}, but that the
imaginary part of \emph{Kong241} seems to behave considerably better for most
part than its real counterpart. To compare the real and the imaginary parts is
insofar interesting as the digital filters are purely real valued.

@@

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker, GEO.17.Werthmuller}; the model parameters are given in
subplot (a). The filter used for this example for the Hankel transform is a
2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for the
inversion and the check of quality. Vertical source and receiver separation is
1\,m, resistivity $\rho=200\,\Omega\,$m, $\varepsilon_\mr{r}$=10,
$\mu_\mr{r}=1$. For the Fourier transform a 4096\,pt FFT was used with
regularly spaced frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up
to 2048\,MHz for both the calculation with the adaptive quadrature (QUA) and
with the DLF. The frequency result is multiplied with a Ricker wavelet with a
center frequency of 250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is
applied. The calculation with the digital filter took under 9 minutes and is
therefore roughly 80 times faster than the adaptive quadrature calculation
which took roughly 11 hours and 27 minutes. Calculating the same model with QWE
took 7 hours and 20 minutes. However, QWE uses in roughly 1/3 of the
calculation internally the adaptive quadrature in this example, see
\cite{GEO.17.Werthmuller}. (Note that DLF was run in parallel using 4 threads
at once, taking effectively only 2 minutes and 10 seconds to calculate.
%EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
%minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.)
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QAGSE from the Fortran
  QUADPACK library) or with (c) a 2001\,pt digital linear filter (b). The
  quadrature took about 11\,h 27\,min to calculate, whereas the filter took
  2\,min 10\,s in parallel on 4 nodes, hence under 9\,min in total.}
%

Figure~\ref{fig:ftGPR} shows in (a)-(c) the real part of the frequency domain
results and in (d)-(f) the time domain results for offsets of 0.2\,m, 2.0\,m,
and 3.0\,m.
%
\plot*{ftGPR}{width=\textwidth}{The real-part frequency-domain and the
time-domain responses for offsets of (a/d) 0.2\,m, (b/e) 2.0\,m, and (c/f)
3.0\,m.}
%

The 2001\,pt filter was derived with the fullspace solution of a medium which
is very similar to the layer in which source and receiver reside in our GPR
example. To show that this filter can also be applied to different layer
parameters we run a test where we swapped layers one and two of the model
in Figure \ref{fig:GPR} (a). The resulting real and imaginary parts for the
offsets 0.2\,m, 2.0\,m, and 3.0\,m are given in Figure~\ref{fig:ftGPR2nd}.

%
\plot*{ftGPR2nd}{width=\textwidth}{The real and imaginary parts of the
  frequency-domain response for the model given in Figure~\ref{fig:GPR} (a)
  with swapped first and second layers for offsets of (a) 0.2\,m, (b) 2.0\,m,
  and (c) 3.0\,m.}
%

These examples show clearly that the filter method can indeed be applied to
high frequency EM modeling and therefore wave phenomena. We are convinced that
with further tests and analysis much better filters could be achieved, and
various concepts could be checked. One approach is to derive a filter for each
frequency band, say one for 1\,MHz--10\,MHz, one for 10\,MHz--100\,MHz, and one
for 100\,MHz--1\,GHz. Another approach could be to derive distinct filters for
$J_0$ and $J_1$, with different spacing and shift values. The first idea would
triple the calculation cost, the second idea double them; however, they both
would still be very fast compared to standard quadrature methods. One could
also apply the digital filter method to acoustic and elastic wavefields as
well.

@@

The presented, free and open-source code \texttt{fdesign} can be used to design
digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}. The code is
available from GitHub in the add-on repository of \texttt{empymod}, called
\texttt{empyscripts}, from version 0.1.2 onwards.

@@

We except the presented algorithm is useful for at least 3 scenarios:
\begin{enumerate}
  \item Provide a fast method to design problem-specific filters. For bigger
    inversion projects (such as, for instance, massive stochastic CSEM
    inversions) a purpose designed, short filter might save much more time than
    it costs to design it. This could even be integrated into inversion codes,
    as an optional pre-inversion step.
  \item Extend the filter method to new areas, namely higher frequencies and
    acoustic and elastic wavefields.
  \item Raise a new interest for digital linear filters in geophysics, by
    making it very easy for anyone to play around, create their own filters,
    and get a better understanding of it. We are sure there are many great
    filters to be discovered.
\end{enumerate}

\section{Acknowledgment}
D.W. would like to thank his colleague Oleg Gurevich Titov, who introduced him
to linear digital filters in the first place, and Jürg Hunziker for valuable
feedback which improved this manuscript considerably.
% This project would not have been successful without the fruitful discussions
% with Evert Slob and Kerry Key. Many thanks as well to Kerry Key for providing
% me with his Matlab scripts, which served as a starting point and initiated
% this project.

% REFERENCES
\begin{thebibliography}{}
\itemsep0pt

\bibitem[Anderson, 1973]{USGS.73.Anderson}
Anderson, W.~L.,  1973, Fortran {IV} programs for the determination of the
  transient tangential electric field and vertical magnetic field about a
  vertical magnetic dipole for an m-layer stratified earth by numerical
  integration and digital linear filtering: USGS, {\bf PB221240}.

\bibitem[Anderson, 1975]{USGS.75.Anderson}
--------, 1975, Improved digital filters for evaluating {F}ourier and {H}ankel
  transform integrals: USGS, {\bf PB242800}.
\newblock
  (\href{https://pubs.er.usgs.gov/publication/70045426}{https://pubs.er.usgs.gov/publication/70045426}).

\bibitem[Anderson, 1979]{GEO.79.Anderson}
--------, 1979, Numerical integration of related {H}ankel transforms of orders
  0 and 1 by adaptive digital filtering: Geophysics, {\bf 44}, 1287--1305.
\newblock (\href{http://doi.org/10.1190/1.1441007}{doi: 10.1190/1.1441007}).
